Implementation of a stable, high-power optical lattice for quantum gas microscopy

### 2018-09-25 LatticeOptics/submission_rev_sci_inst/latticeoptics.tex
rev sci inst submission reviewer 1 addressed

\begin{figure*}
  \begin{center}
    \includegraphics{Figure2.pdf}
    \caption{\textbf{Laser noise of the lattice system.} (a) Shows the laser noise of unmodified and free-running Nufern fiber amplifiers and the Mephisto seed, as well as an independent Mephisto MOPA 55W (not used for the optical lattice presented in this manuscript). (b) Shows the open-loop and closed-loop behavior of the high-power feedback loop using the  modified fiber amplifiers, as described in the main text. (c) Shows the open-loop behavior of the low-power feedback loop using the modified fiber amplifiers, including an additional low-pass filtering stage to limit the loop bandwidth. (d) Closed-loop behavior of the low-power feedback loop without additional low-pass filtering, measured on the same setup as panel (c).}\label{fig:noises}
  \end{center}
\end{figure*}

@@

The NuAmp suffers from several technical problems.
It is controlled via a USB interface, which complicates communications with an array of these devices.
Worse yet, it is suspected of causing ground loops which add noise and interference to the analog systems.
This added noise can be seen in Fig.~\ref{fig:noises}, where the fiber-amplified light exhibits many more spurs in the intensity noise spectrum than the seed laser.
We solve this problem by replacing the USB control board with a custom-designed solution  that is compatible with our laboratory control system and engineered to avoid ground loops.

@@

The large-diameter fiber tip at the output of the fiber amplifier is mounted in a monolithic mount made from copper.
The beam is collimated using an $f=\unit{20}{mm}$ fused-silica, air-spaced triplet collimator (Opto-Sigma HFTLSQ-15-20PF1) angled with respect to and centered on the angle-cleaved fiber tip.
The polarization is set using an IBS coated thin-film plate polarizer (Precision photonics PO1056-FY), which has the added benefit of rejecting the undesired cladding modes of the fiber, which carry about \unit{1}{W} of light.
The fact that the lattice is retro-reflected, combined with the observation that optical isolation falls with applied optical power~\cite{Yoshida1999} means that two stages of optical isolators must be used.
The optical isolators are based on \unit{5}{mm} aperture Tb$_3$Ga$_5$O$_12$ (TGG) isolators (Thorlabs IO-5-1064-VHP), where we replaced the default polarizers (BK7 Brewster plates) by IBS-coated fused silica plate polarizers.
The isolators are located approximately \unit{1}{m} away from the ultracold atoms which are susceptible to stray magnetic fields.
To overcome this problem the isolators are enclosed in mu-metal shielding.
Actuation of the power is performed after the optical isolators, such that the full fiber amplifier output continuously passes through them, minimizing thermal transients during the experimental sequence.

@@

After the isolation stage, we use the sequence of optical elements shown in Fig.~\ref{fig:optical_layout}, consisting of a polarizer, half-wave plate (HWP), Berek compensator mounted on a precision galvo (Thorlabs GVS002, or Camtech 8320K, depending on the beam path) half-wave plate, and polarizer.
The Berek compensator is a $z$-cut quartz plate anti-reflection-coated at \unit{1064}{nm}.
By tilting the plate about its vertical axis by a few degrees, the extraordinary axis is mixed into the propagation of the beam, leading to tilt-dependent birefringence that can be tuned from a zero-wave plate past a half-wave plate, represented by a retardation angle $\phi$.
Combined with the subsequent polarization optics, rotation of the Berek compensator can vary the transmitted power with a contrast of $\sim$1000:1.
Figure~\ref{fig:berek} shows the transmitted power as a function of retardation angle $\phi$ and fast-axis angle $\theta$ of the waveplates surrounding the compensator.
\begin{figure}
  \begin{center}
    \includegraphics{Figure4.pdf}
    \caption{\textbf{Berek compensator} transmission as a function of the Berek compensator's retardation angle $\phi$ and waveplate rotation angles $\theta$.}\label{fig:berek}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{Figure5.pdf}
    \caption{\textbf{Berek compensator based feedback loop}. This Berek compensator based feedback loop is used to vary the beam intensity at relatively high power. (a) Shows the small signal impulse response and (b-c) show the corresponding Bode plots of the transfer function $G$.}\label{fig:berek_step_response}
  \end{center}
\end{figure}
A key point, as seen in Fig.~\ref{fig:berek}, is that if the waveplates are rotated together, the maximum transmitted power is fixed, while the minimum can be raised.
The waveplates are detuned such that the maximum power needed in \emph{interaction} mode is the minimum power transmitted by the Berek compensator system, strongly suppressing fluctuations due to angle positioning noise of the Berek compensator.
Naturally, rotation of a parallel plate in the beam path induces a shift of the beam, but this is acceptable because the Berek compensator is only active during the imaging phase of the experiment, when the position of the underlying harmonic trap is largely irrelevant.
Fortunately, the spatial phase of the optical lattice is set by the retro-reflector mirror, which does not move, and is not affected by the rotation of the Berek compensator.
This system functions as the actuator arm of a digital feedback loop, where the power of the beam is measured using a low noise photodiode. This feedback loop regulates the lattice between \unit{1}{W} of optical power and its maximum value (which depends on the amplifier, but is $\approx$\unit{20}{W}).
The step response and corresponding Bode plots~\cite{Bechhoefer2005} are shown in Fig.~\ref{fig:berek_step_response}.
A polarization rotation by $2 \pi$ is produced by approximately $3^\circ$ of axial rotation of the quartz plate.
Since we never need to rotate the plate outside this range, the position is constrained to this window in software.

Because two feedback loops are present in the system (the second one detailed in the next section), care must be taken so that they do not function simultaneously.
To accomplish this, the low-power loop is always manually railed to its high extreme when the high-power loop is active, and the digital high-power loop is disabled when the low-power loop is active.

@@

The connection between the actuation and monitoring mechanisms of the feedback system is the loop filter, which compares the \emph{set-point} signal, $V_\mathrm{SP}(t)$, supplied by the experimental control system and the photodiode signals $V_\mathrm{PD}(t)$, to adjust the actuation accordingly.
We use a simple, purely integrating design based in three stages, as seen in Fig.~\ref{fig:circuits}(c), based on a low-noise op-amp (Texas Instruments OPA211).
The first stage computes and amplifies an error signal, $e(t)=V_\mathrm{SP}(t)-V_\mathrm{PD}(t)$, which is integrated by the second stage.
The integral gain is set using a low-drift trim-pot.
This is the case because within relevant frequency ranges, the AOM can be coarsely approximated by a pure time delay process, given by the acoustic wave transit time of $\approx$\unit{1}{\mu{}s} across the active region of the AOM crystal.
In this scenario, the proportional term must be small to avoid instability at the $\pi$ phase shift point, and a pure integral controller is optimal~\cite{Skogestad2003}.
For debugging purposes, the loop filter features buffered photodiode monitor and error signal monitor outputs.

The third stage of the loop filter is used to accommodate the fact that the actuation arm is nonlinear: the optical power transmitted by the AOM is proportional to the RF power incident on it $P_{\mathrm{rf}}\propto V_{IF}^2$, where $V_{IF}$ is the actuation signal supplied to the IF port of the mixer.
In effect, if the rest of the feedback system was linear, this would correspond to a quadratically increasing gain until the AOM is saturated, which could lead to slow behavior at low signal levels and ringing at higher signal levels.
Even worse, since the mixer is insensitive to the sign of the voltage applied, the gain is becomes negative for negative voltages, resulting in an unstable system.

@@

Generally, changes in environmental conditions (such as temperature and relative humidity), load on the water-cooling systems and other external conditions can lead to a slow drift in the laser pointing over long timescales.
The situation is further complicated by the fact that opening the enclosure of the experiment changes the thermal state of the system.
For these reasons, each lattice axis is fitted with a remotely actuated mirror, controlled by the experimental control software over the lab network, allowing for non-invasive, digital alignment of the system.
The remote actuators are ``Picomotors'' (Newport 8302), fitted into a customized \unit{2}{inch} mirror mounts.
They work in a ``fire-and-forget'' manner, remaining passively stable when not in active use.

@@

In our experiments, the position of the Gaussian beam envelope of the optical lattice is impossible to determine from the site-resolved image of the atoms, because the atoms may not occupy the central part of the envelope depending on the specefics of the lattice loading procedure.
In addition, the atomic distributions are governed by the sum of the incident lattices and thus cannot be used to reliably infer the degree of overlap between the lattice beams.
Nevertheless, it is possible to adiabatically load a strongly anisotropic optical lattice by lowering the power of one of the lattice beams, resulting in an atomic distribution that closely matches the trap produced by the high-power beam~\cite{Parsons2015}.
This procedure yields the correct position of the Gaussian beam envelope in the atom plane, which we can use to analyze each beam's passive stability and its relationship to the beam-monitoring camera.

@@

For this reason, we developed a procedure to periodically update the affine transform relating the beam-monitoring camera to the atom plane.
We test this procedure, we analyze the data set of monitoring-camera measurements and atomic positions (shown in Fig.~\ref{fig:stability}) in the following way: we split the data set into sequential blocks of 20 shots, each followed by a block of 8 shots.
We try to predict the larger blocks, by analyzing the smaller blocks to derive the updates to the affine transform.
The results are shown in Fig.~\ref{fig:stability}(d-f) and display a large degree of tracking between the predicted and measured atom positions.
The average taxicab distance between the predicted position and the measured position is 0.7~lattice sites.

>>>>

\begin{figure*}
  \begin{center}
    \includegraphics{Figure2.pdf}
    \caption{\textbf{Laser noise of the lattice system.} (a) Shows the laser noise of unmodified and free-running Nufern fiber amplifiers and the Mephisto seed, as well as an independent Mephisto MOPA 55W (not used for the optical lattice presented in this manuscript). (b) Shows the open-loop and closed-loop behavior of the high-power feedback loop using the  modified fiber amplifiers, as described in the main text. (c) Shows the open-loop behavior of the low-power feedback loop using the modified fiber amplifiers, including an additional low-pass filtering stage to limit the loop bandwidth at various constant powers. (d) Closed-loop behavior of the low-power feedback loop without additional low-pass filtering, measured on the same setup as panel (c).}\label{fig:noises}
  \end{center}
\end{figure*}

@@

The NuAmp suffers from several technical problems.
It is controlled via a USB interface, which complicates communications with an array of these devices.
Worse yet, it is suspected of causing ground loops which add noise and interference to the analog systems.
This added noise can be seen in Fig.~\ref{fig:noises}, where the fiber-amplified light exhibits many more spurs in the intensity noise spectrum than the seed laser, particularly near \unit{10}{kHz}.
We solve this problem by replacing the USB control board with a custom-designed solution  that is compatible with our laboratory control system and engineered to avoid ground loops.

@@

The large-diameter fiber tip at the output of the fiber amplifier is mounted in a monolithic mount made from copper.
The beam is collimated using an $f=\unit{20}{mm}$ fused-silica, air-spaced triplet collimator (Opto-Sigma HFTLSQ-15-20PF1) angled with respect to and centered on the angle-cleaved fiber tip.
The polarization is set using an IBS coated thin-film plate polarizer (Precision photonics PO1056-FY), with the added benefit of rejecting the undesired cladding modes of the fiber, which carry about \unit{1}{W} of light. A thin-film polarizer was used to minimize the number of normal and backreflecting surfaces near the output of the fiber in addition to providing excellent polarization performance.
The fact that the lattice is retro-reflected, combined with the observation that optical isolation falls with applied optical power~\cite{Yoshida1999} means that two stages of optical isolators must be used.
The optical isolators are based on \unit{5}{mm} aperture Tb$_3$Ga$_5$O$_12$ (TGG) isolators (Thorlabs IO-5-1064-VHP), where we replaced the default polarizers (BK7 Brewster plates) by IBS-coated fused silica plate polarizers.
The isolators are located approximately \unit{1}{m} away from the ultracold atoms which are susceptible to stray magnetic fields.
To overcome this problem the isolators are enclosed in mu-metal shielding.
Actuation of the power is performed after the optical isolators, such that the full fiber amplifier output continuously passes through them, minimizing thermal transients during the experimental sequence.

@@

After the isolation stage, we use the sequence of optical elements shown in Fig.~\ref{fig:optical_layout}, consisting of a polarizer, half-wave plate (HWP), Berek compensator mounted on a precision galvo (Thorlabs GVS002, or Camtech 8320K, depending on the beam path), half-wave plate, and polarizer.
The Berek compensator is a $z$-cut quartz plate anti-reflection-coated at \unit{1064}{nm}.
By tilting the plate about its vertical axis by a few degrees, the extraordinary axis is mixed into the propagation of the beam, leading to tilt-dependent birefringence that can be tuned from a zero-wave plate past a half-wave plate, represented by a retardation angle $\phi$.
Combined with the subsequent polarization optics, rotation of the Berek compensator can vary the transmitted power with a contrast of $\sim$1000:1.
Figure~\ref{fig:berek} shows the transmitted power as a function of retardation angle $\phi$ and fast-axis angle $\theta$ of the waveplates surrounding the compensator.
\begin{figure}
  \begin{center}
    \includegraphics{Figure4.pdf}
    \caption{\textbf{Berek compensator} transmission as a function of the Berek compensator's retardation angle $\phi$ and waveplate rotation angles $\theta$.}\label{fig:berek}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{Figure5.pdf}
    \caption{\textbf{Berek compensator based feedback loop}. This Berek compensator based feedback loop is used to vary the beam intensity at relatively high power. (a) Shows the small signal impulse response of the feedback loop (step size of $\approx 2$ W) and (b-c) show the corresponding Bode plots of the closed loop transfer function $G$.}\label{fig:berek_step_response}
  \end{center}
\end{figure}
A key point, as seen in Fig.~\ref{fig:berek}, is that if the waveplates are rotated together, the maximum transmitted power is fixed, while the minimum can be raised.
The waveplates are detuned such that the maximum power needed in \emph{interaction} mode is the minimum power transmitted by the Berek compensator system, strongly suppressing fluctuations due to angle positioning noise of the Berek compensator, since the slope of transmission with respect to the Berek angle is $0$ at this point.
Naturally, rotation of a parallel plate in the beam path induces a shift of the beam, but this is acceptable because the Berek compensator is only active during the imaging phase of the experiment, when the position of the underlying harmonic trap is largely irrelevant.
Fortunately, the spatial phase of the optical lattice is set by the retro-reflector mirror, which does not move, and is not affected by the rotation of the Berek compensator.
This system functions as the actuator arm of a digital feedback loop, where the power of the beam is measured using a low noise photodiode. This feedback loop regulates the lattice between \unit{1}{W} of optical power and its maximum value (which depends on the amplifier, but is $\approx$\unit{20}{W}).
The step response and corresponding Bode plots~\cite{Bechhoefer2005} are shown in Fig.~\ref{fig:berek_step_response}.
A polarization rotation by $2 \pi$ is produced by approximately $3^\circ$ of axial rotation of the quartz plate.
Since we never need to rotate the plate outside this range, the position is constrained to this window in software.

Because two feedback loops are present in the system (the second one detailed in the next section), care must be taken so that they do not function simultaneously.
To accomplish this, the low-power loop is always manually railed to its high extreme when the high-power loop is active.
This is done by shuttering the photodiode and directing the set-point to a high value, which quickly winds up the integrator to its maximum output.
The digital high-power loop is disabled in software when the low-power loop is active.

@@

The connection between the actuation and monitoring mechanisms of the feedback system is the loop filter, which compares the \emph{set-point} signal, $V_\mathrm{SP}(t)$, supplied by the experimental control system and the photodiode signals $V_\mathrm{PD}(t)$, to adjust the actuation accordingly.
We use a simple, purely integrating design in three stages, as seen in Fig.~\ref{fig:circuits}(c), based on a low-noise op-amp (Texas Instruments OPA211).
The first stage computes and amplifies an error signal, $e(t)=V_\mathrm{SP}(t)-V_\mathrm{PD}(t)$, which is integrated by the second stage.
The integral gain is set using a low-drift trim-pot.
This is the case because within relevant frequency ranges, the AOM can be coarsely approximated by a pure time delay process, given by the acoustic wave transit time of $\approx$\unit{1}{\mu{}s} across the active region of the AOM crystal.
In this scenario, the proportional term must be small to avoid instability at the $\pi$ phase shift point, and a pure integral controller is optimal~\cite{Skogestad2003}.
For debugging purposes, the loop filter features buffered photodiode monitor and error signal monitor outputs.

The third stage of the loop filter is used to accommodate the fact that the actuation arm is nonlinear: the optical power transmitted by the AOM is proportional to the RF power incident on it $P_{\mathrm{rf}}\propto V_{IF}^2$, where $V_{IF}$ is the actuation signal supplied to the IF port of the mixer.
In effect, if the rest of the feedback system was linear, this would correspond to a quadratically increasing gain until the AOM is saturated, which could lead to slow behavior at low signal levels and ringing at higher signal levels.
Even worse, since the mixer is insensitive to the sign of the voltage applied, the gain is negative for negative voltages, resulting in an unstable system.

@@

Generally, changes in environmental conditions (such as temperature and relative humidity), load on the water-cooling systems and other external conditions can lead to a slow drift in the laser pointing over long timescales.
The situation is further complicated by the fact that opening the enclosure of the experiment changes the thermal state of the system.
For these reasons, each lattice axis is fitted with a remotely actuated mirror, controlled by the experimental control software over the lab network, allowing for non-invasive, digital alignment of the system.
The remote actuators are ``Picomotors'' (Newport 8302), fitted into customized \unit{2}{inch} mirror mounts.
They work in a ``fire-and-forget'' manner, remaining passively stable when not in active use.

@@

In our experiments, the position of the Gaussian beam envelope of the optical lattice is impossible to determine from the site-resolved image of the atoms, because the atoms may not occupy the central part of the envelope depending on the specifics of the lattice loading procedure.
In addition, the atomic distributions are governed by the sum of the incident lattices and thus cannot be used to reliably infer the degree of overlap between the lattice beams.
Nevertheless, it is possible to adiabatically load a strongly anisotropic optical lattice by lowering the power of one of the lattice beams, resulting in an atomic distribution that closely matches the trap produced by the high-power beam~\cite{Parsons2015}.
This procedure yields the correct position of the Gaussian beam envelope in the atom plane, which we can use to analyze each beam's passive stability and its relationship to the beam-monitoring camera.

@@

For this reason, we developed a procedure to periodically update the affine transform relating the beam-monitoring camera to the atom plane.
To test this procedure, we analyze the data set of monitoring-camera measurements and atomic positions (shown in Fig.~\ref{fig:stability}) in the following way: we split the data set into sequential blocks of 20 shots, each followed by a block of 8 shots.
We try to predict the larger blocks, by analyzing the smaller blocks to derive the updates to the affine transform.
The results are shown in Fig.~\ref{fig:stability}(d-f) and display a large degree of tracking between the predicted and measured atom positions.
The average taxicab distance between the predicted position and the measured position is 0.7~lattice sites.
This result shows that recalibrating the affine transformation every 20 shots bounds the tracking error to an acceptable level. In practical applications, larger blocks of shots containing scientific measurements are interleaved with small recalibration blocks.
